pirouette: Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-03-15 article.tex
Review ERQ 3, progress #218

\subsection{Twin tree}

@@

@@

"What is the error made by BEAST2 from a phylogeny, 
when hand-picking an inference model, compared to the background noise?"

Most of the settings for this example research question are
the same as in the previous section, which are
the phylogeny from an unknown speciation model (listing~\ref{lst:unknown_phylogeny}), 
the alignment parameters (listing~\ref{lst:create_alignment}) 
and the list of experiments (listing~\ref{lst:create_generative_experiment}).

This time, however, we are interested in creating a twin tree,
as such a twin tree allows us to measure a baseline error produced
by the BEAST2 run.
A twin tree
has the same topology as the given tree, 
yet its branching times are simulated according 
to a tree prior already implemented by BEAST2 to produce a posterior. 
Because for a twin tree we know the tree prior 
and because we use that tree prior in inference, 
we expect BEAST2 to produce a posterior of phylogenies
that is close to the twin tree.
As the inference model for the twin tree is (more) completely
specified than the true tree, 
we expect that a twin tree and its posterior trees are more similar
than a true tree and its posterior trees.
Due to this, the twinning is used a control for the experiment.

Creating a twinning parameter is easy, as it has sensible default settings:

@@

We can now measure the errors made by BEAST2 
when inferring the given phylogeny 
versus the error it makes when the same procedure is applied to the twin tree.

All the \verb;pirouette; arguments are bundled and checked by \verb;create_pir_params;:

@@

Running:

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_3_errors.png}
  \caption{
    Example 3: the error made by BEAST2 from a phylogeny, picking the best inference model versus the control.
    Here, the twin column shows the error BEAST2 makes starting from the original tree versus the twin. 
  }
  \label{fig:example_3}
\end{figure}

\iffalse
The error BEAST2 makes from a phylogeny 
picking the best inference model, compared to the background noise.
Here, the twin column shows the error BEAST2 makes on an idealized
tree, to measure the noise, which is the minimal error. 
\fi

>>>>

\subsection{Twin tree}\label{subsec:twinning}

@@

In ERQ 1 we measured the baseline error of the \verb;pirouette; method, 
by using a standard tree prior in both the generation of the tree 
and in our inference.
The goal of \verb;pirouette; is to quantify the 
influence of a new (non-standard) tree prior.
To do so, we need to compare the inference error 
from the new tree prior to a baseline error,
by use of a twin tree (see subsection~\ref{subsec:twinning}).
In this ERQ, we assume the new tree prior so closely related to
a standard tree prior, that we hand-pick a standard tree prior to
serve as a stand-in generative tree prior.

@@

"What is the error made by BEAST2 from a phylogeny, 
when hand-picking a tree prior, compared to the background noise?"

Most of the settings for this example research question are
the same as in the previous section, which are
the phylogeny from an unknown speciation model (listing~\ref{lst:unknown_phylogeny}), 
the alignment parameters (listing~\ref{lst:create_alignment})
and the list of experiments (listing~\ref{lst:create_generative_experiment}),
the latter stating that the generative tree prior is a Yule tree prior.

This time, however, we are interested measuring both the inference error
and a baseline error, which is achieved by creating a twinning parameter
set. Doing so is easy, as it has sensible default settings:

@@

Bundling all the \verb;pirouette; arguments by \verb;create_pir_params;:

@@

Running and plotting the results:

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
errors <- pir_run(
  phylogeny,
  pir_params = pir_params
)
pir_plot(errors)
\end{lstlisting}

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_3_errors.png}
  \caption{
    Example 3: the error made by BEAST2 from a phylogeny, 
    when hand-picking a generative tree prior versus the control.
    Here, the 'twin' column shows the baseline inference error,
    as measured on the twin tree
  }
  \label{fig:example_3}
\end{figure}
